---
title: "activity_analysis"
author: "Ben Hicks"
date: "15/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F)

# Packages
suppressPackageStartupMessages(library(tidygraph))
suppressPackageStartupMessages(library(ggraph))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lakit))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(TDAmapper))
suppressPackageStartupMessages(library(TDA))
suppressPackageStartupMessages(library(viridis))
suppressPackageStartupMessages(library(colorspace))
suppressPackageStartupMessages(library(patchwork))


# Data
prep_aa <- function(aa) {
    aa %>% 
        as_tibble() %>% 
        select(person, subject, subject_code, role, row_status,
               session, event, handle, data, content_pk1, content, 
               grade_quartile, starts_with("cluster"),
               timestamp)
}

# Functions
aa_to_person_intervals <- function(aa, cluster_var, frac = 0.01) {
    df <- aa %>% 
        rename(cluster = {{cluster_var}}) %>% 
        group_by(person, session, cluster,
                 grade_quartile, row_status) %>% 
        nest() %>% 
        mutate(nums = map_int(data, nrow)) %>% 
        filter(nums > 1) %>% 
        select(-nums) %>% 
        mutate(interval_dfs = map(
            data, 
            ~data.frame(
                intervals = 
                    as.numeric(
                        timelist_to_difference(.$timestamp, frac = frac))
                ))) %>%
        select(-data) %>% 
        unnest(cols = interval_dfs) %>% 
        mutate(intervals = as.duration(intervals)) %>% 
        filter(as.duration(intervals) < as.duration('1 day'),
               !is.na(grade_quartile))
    df$cluster <- factor(df$cluster)
    return(df)
}

aa_to_person_intervals_all_clusters <- function(aa, frac = 0.05) {
    df <- aa %>% 
        group_by(person, session, 
                 cluster_2, cluster_3, cluster_4, cluster_5,
                 grade_quartile, row_status) %>% 
        nest() %>% 
        mutate(nums = map_int(data, nrow)) %>% 
        filter(nums > 1) %>% 
        select(-nums) %>% 
        mutate(interval_dfs = map(
            data, 
            ~data.frame(
                intervals = 
                    as.numeric(
                        timelist_to_difference(.$timestamp, frac = frac))
                ))) %>%
        select(-data) %>% 
        unnest(cols = interval_dfs) %>% 
        mutate(intervals = as.duration(intervals)) %>% 
        filter(as.duration(intervals) < as.duration('1 day'),
               !is.na(grade_quartile))
    df$cluster_2 <- factor(df$cluster_2)
    df$cluster_3 <- factor(df$cluster_3)
    df$cluster_4 <- factor(df$cluster_4)
    df$cluster_5 <- factor(df$cluster_5)
    return(df)
}

plot_clusters <- function(n, cluster_var) {
    DF$nd[[n]] %>% 
        filter(Person != "Student321") %>% 
        ggplot(aes(x = Dim1, y = Dim2, color = factor({{cluster_var}}))) +
        geom_point() +
        theme_minimal() +
        theme(legend.position = "none") +
        scale_color_discrete_qualitative(palette = "Dark 3") +
        geom_mark_ellipse(aes(label = {{cluster_var}}))
}
```

``` {r DF_setup, eval = F}
# Setting up DF
DF <- tibble(
    subject = c("Bio", "Bio", "Soc", "Soc"),
    year = c(2017, 2018, 2017, 2018),
    aa = list(rdata.psych::aaBio17 %>% prep_aa(),
              rdata.psych::aaBio18 %>% prep_aa(),
              rdata.psych::aaSoc17 %>% prep_aa(),
              rdata.psych::aaSoc18 %>% prep_aa()),
    nd = list(rdata.psych::nodesBio17 %>% drop_na() %>% as_tibble(),
              rdata.psych::nodesBio18 %>% drop_na() %>% as_tibble(),
              rdata.psych::nodesSoc17 %>% drop_na() %>% as_tibble(),
              rdata.psych::nodesSoc18 %>% drop_na() %>% as_tibble())
)
```


``` {r building_interval_dfs, cache = TRUE, eval = FALSE}
# Building intervals dfs
intervals_A1 <- aaA1 %>% 
    aa_to_person_intervals_all_clusters()
intervals_A2 <- aaA2 %>% 
    aa_to_person_intervals_all_clusters()
intervals_B1 <- aaB1 %>% 
    aa_to_person_intervals_all_clusters()
intervals_B2 <- aaB2 %>% 
    aa_to_person_intervals_all_clusters()
```

``` {r writing_interval_dfs, eval = FALSE}
write_rds(intervals_A1, 'intervals_A1_stored.rds')
write_rds(intervals_A2, 'intervals_A2_stored.rds')
write_rds(intervals_B1, 'intervals_B1_stored.rds')
write_rds(intervals_B2, 'intervals_B2_stored.rds')
```

``` {r readings_interval_dfs, eval = FALSE}
intervals_A1 <- read_rds('intervals_A1_stored.rds') # B-17
intervals_A2 <- read_rds('intervals_A2_stored.rds')
intervals_B1 <- read_rds('intervals_B1_stored.rds')
intervals_B2 <- read_rds('intervals_B2_stored.rds')
# TODO : Re-write sampling of intervals but per student (this will help with comparing with TDA later)
# To speed up sample the timestamps; 100 starts and 100 finishes, with replacement, then take the difference of the 
# two vectors.
# Actually - might just need to join the new groups by person id.
```

``` {r adding_interval_dfs_to_DF, eval = F}
DF$int <- list(intervals_B1, intervals_B2, intervals_A1, intervals_A2)
rm(intervals_B1, intervals_B2, intervals_A1, intervals_A2)
```

# Topological Data Analysis

``` {r tda_functions}
# TDA; Mapper, Persistent Homology
nodes_to_matrix <- function(nodes, scale = FALSE, marks = FALSE) {
    nodes <- nodes %>% filter(complete.cases(nodes)) # shouldn't be needed as NA dropped earlier
    X <- nodes %>% 
        select(Grade, Accesses, 
               mean_clicks_per_access, sd_clicks, 
               median_time_per_access, sd_time, TotalTime, 
               ForumViews)
    if (!marks) {X <- X %>% select(-Grade)}
    if (scale) {X <- scale(X)}
    return(X)
}

as_duration_approx <- function(s) {
    d <- as.duration(s) %>% 
        as.character() %>% 
        str_remove("^.* \\(") %>% 
        str_remove("\\)$")
    return(d)
}

nodes_text_summary <- function(nodes) {
    n <- nrow(nodes)
    grade <- mean(nodes$Grade, na.rm = T) %>% round(3)
    accesses <- mean(nodes$Accesses, na.rm = T) %>% round(0)
    clicks_pa <- mean(nodes$mean_clicks_per_access, na.rm = T) %>% round(0)
    clicks_sd <- mean(nodes$sd_clicks, na.rm = T) %>% round(0)
    time_md <- mean(nodes$median_time_per_access, na.rm = ) 
    time_md_print <- time_md %>% 
        round(0) %>% 
        as_duration_approx()
    time_sd <- mean(nodes$sd_time, na.rm = T) 
    time_sd_print <- time_sd %>% 
        round(0) %>% 
        as_duration_approx()
    time <- mean(nodes$TotalTime, na.rm = T) 
    time_print <- time %>% 
        round(0) %>% 
        as_duration_approx()
    forum <- mean(nodes$ForumViews, na.rm = T) %>% round(0)
    text_out <- paste0("n: ", n, " G: ", grade, ", F: ", forum,
                       "\nA: ", accesses, ":", clicks_pa, ":", clicks_sd,
                       "\nT: ", time_print, ":", time_md_print, ":", time_sd_print)
    df_out <- data.frame(summary = text_out,
                         n = n, 
                         grade = grade,
                         forum = forum,
                         accesses = accesses,
                         clicks_pa = clicks_pa,
                         clicks_sd = clicks_sd,
                         time = time,
                         time_md = time_md,
                         time_sd = time_sd)
    return(df_out)
}

mapper_vertex_summaries <- function(M, nodes) {
    n_nodes <- length(M$points_in_vertex)
    df <- 1:n_nodes %>% 
        as.list() %>% 
        map_df(~nodes_text_summary(nodes[M$points_in_vertex[[.x]] , ]))
    df <- df %>% 
        mutate(id = row_number()) %>% 
        select(id, everything())
    return(as_tibble(df))
}

cluster_vertex_summaries <- function(nodes, cluster_var) {
    n_clusters <- nodes %>% 
        pull({{cluster_var}}) %>% 
        unique() %>% 
        length()
    df <- 1:n_clusters %>% 
        as.list() %>% 
        map_df(~nodes_text_summary(nodes %>% filter({{cluster_var}} == .x)))
    df <- df %>% 
        mutate(id = row_number()) %>% 
        select(id, everything())
    return(as_tibble(df))
}

plot_mapper_d3 <- function(M, nodes, ...) {
    nodes <- nodes %>% filter(complete.cases(nodes))
    v <- mapperVertices(M, nodes$Person)
    e <- mapperEdges(M)
    return(networkD3::forceNetwork(Links = e, 
                                   Nodes = v,
                                   Source = "Linksource", 
                                   Target = "Linktarget",
                                   Value = "Linkvalue",
                                   Group = "Nodegroup",
                                   NodeID = "Nodename",
                                   Nodesize = "Nodesize",
                                   ...))
}


plot_nodes_through_mapper_d3 <- function(nodes) {
    plot_mapper_d3(nds_to_mapper(nodes), nodes)
}

nodes_to_mapper <- function(nodes, scale = TRUE, marks = FALSE,
                          fv = NULL,
                          ni = 6, po = 50, nb = 10) {
    X <- nodes_to_matrix(nodes, scale = scale, marks = marks)
    dist <- dist(X)
    if (is.null(fv)) {
        fv <- X[, 4]
    } else {
        fv <- X %>% as_tibble() %>% pull({{fv}})
    }
    Mapper <- mapper(dist_object = dist,
                     filter_values = fv,
                     num_intervals = ni,
                     percent_overlap = po,
                     num_bins_when_clustering = nb)
    return(Mapper)
}

mapper_and_nodes_to_plot <- function(mapper, nodes, lyt = "fr", col_var, use_labels = FALSE, remove_isolates = F,
                                     color_option = "D") {
    # Takes a mapper and the original nodes and creates a network plot with summary statistics
    mapper_nodes <- mapper_vertex_summaries(mapper, nodes) 
    edges <- as_tbl_graph(mapper$adjacency) %>%
        activate(edges) %>% 
        as_tibble()
    graph <- tbl_graph(nodes = mapper_nodes,
                       edges = edges)
    if (remove_isolates) {
        graph <- graph %>% 
            activate(nodes) %>% 
            filter(!node_is_isolated())
    }
    g <- graph %>% 
        ggraph(layout = lyt) +
        geom_edge_link(alpha = 0.8) +
        geom_node_point(aes(size = n, color = {{col_var}}), alpha = 0.8) +
        scale_color_viridis(option = color_option) +
        scale_size_continuous(range = c(2,6)) +
        theme_graph() +
        theme(legend.position = "none",
              panel.background = element_rect(fill = "slategray1"))
    if (use_labels) {
        g <- g +
            geom_node_label(aes(label = ifelse(n > 2, paste("Group:", id, summary), id), color = {{col_var}}),
                            repel = F, size = 3) 
    } else {
        g <- g +
            geom_node_text(aes(label = ifelse(n > 2, paste("Group:", id, summary), NA)),
                           repel = F, size = 2) 
    }
    return(g)
}

plot_mapper <- function(n, ...) {
    mapper_and_nodes_to_plot(
        mapper = DF$Mapper[[n]],
        nodes = DF$nd[[n]],
        col_var = grade,
        ...
    )
}

plot_mapper_time <- function(n, ...) {
    mapper_and_nodes_to_plot(
        mapper = DF$Mapper.TotalTime[[n]],
        nodes = DF$nd[[n]],
        col_var = time,
        ...
    )
}

plot_mapper_time_g <- function(n) {
    mapper_and_nodes_to_plot(
        mapper = DF$Mapper.TotalTime.Grade[[n]],
        nodes = DF$nd[[n]],
        col_var = time
    )
}
plot_mapper_time_sd <- function(n) {
    mapper_and_nodes_to_plot(
        mapper = DF$Mapper.sd_time[[n]],
        nodes = DF$nd[[n]],
        col_var = time
    )
}
plot_mapper_time_sd_g <- function(n) {
    mapper_and_nodes_to_plot(
        mapper = DF$Mapper.sd_time.Grade[[n]],
        nodes = DF$nd[[n]],
        col_var = time
    )
}

limits_from_matrix <- function(X) {
    matrix(
        c(apply(X, 2, min),
          apply(X, 2, max))
        , nrow = 2,
        byrow = TRUE)
}

gDiag <- function(X, n = NULL, d = 2) {
    if (!is.null(n)) {
        d <- dim(X)[[1]]
        n <- min(n, d)
        X <- X[sample(1:d, n), ]
    }
    gridDiag(X,
             FUN = distFct,
             lim = limits_from_matrix(X),
             by = 0.3,
             maxdimension = d,
             printProgress = TRUE)
}

plot_set_compare <- function(d, var1, var2) {
    d %>% 
        ggplot(aes(x = factor({{var1}}), 
                   y = factor({{var2}}), 
                   fill = n)) +
        geom_tile() +
        scale_fill_viridis(option = "D") +
        theme_minimal() +
        theme(legend.position = "none")
}
```

``` {r tda, eval = F}
DF <- DF %>% 
    mutate(
        M = map(nd, ~nodes_to_matrix(., scale = TRUE, marks = TRUE)),
        M.no.grade = map(nd, ~nodes_to_matrix(., scale = TRUE, marks = FALSE)),
        X = map(nd, ~scale(.x %>% select(Dim1, Dim2, Dim3))),
        X.g = map(nd, ~scale(.x %>% select(Dim1, Dim2, Dim3, Grade)))
    )

DF <- DF %>% 
    mutate(
        Mapper = map(nd, ~nodes_to_mapper(., fv = "Grade", marks = TRUE)),
        Mapper.TotalTime = map(nd, ~nodes_to_mapper(., fv = "TotalTime")),
        Mapper.TotalTime.Grade = map(nd, ~nodes_to_mapper(., marks = TRUE, fv = "TotalTime")),
        Mapper.sd_time = map(nd, ~nodes_to_mapper(., fv = "sd_time")),
        Mapper.sd_time.Grade = map(nd, ~nodes_to_mapper(., marks = TRUE, fv = "sd_time")),
        Mapper.Accesses = map(nd, ~nodes_to_mapper(., fv = "Accesses")),
        Mapper.Accesses.Grade = map(nd, ~nodes_to_mapper(., marks = TRUE, fv = "Accesses"))
    )

for (n in 1:4) {
    m_max <- length(DF$Mapper[[n]]$points_in_vertex)
    
    DF$nd[[n]]$mapper_vertex <- NA_integer_
    for (m in 1:m_max) {
        DF$nd[[n]][
            DF$Mapper[[n]]$points_in_vertex[[m]],
            ]$mapper_vertex <- m
    }
}

for (n in 1:4) {
    m_max <- length(DF$Mapper.TotalTime[[n]]$points_in_vertex)
    
    DF$nd[[n]]$mapper_vertex_time <- NA_integer_
    for (m in 1:m_max) {
        DF$nd[[n]][
            DF$Mapper.TotalTime[[n]]$points_in_vertex[[m]],
            ]$mapper_vertex_time <- m
    }
}

rm(m, m_max, n)

DF <- DF %>% 
    mutate(
        int = map2(.x = int,
                   .y = nd,
                   ~inner_join(.x, .y %>% select(person = Person, mapper_vertex),
                               by = "person"))
    )

DF <- DF %>% 
    mutate(cluster_mapper_compare = nd %>% 
               map(~.x %>% 
                       group_by(mapper_vertex, cluster_2, cluster_3, cluster_4, cluster_5) %>% 
                       tally()))



# Persistant homology

DF <- DF %>% 
    mutate(pHomDiag = map(X, gDiag),
           pHomDiagGrade = map(X.g, gDiag))

# Other clusters
append_pamk_clusters <- function(nd, M) {
    clustering <- fpc::pamk(M)
    nd <- bind_cols(nd, 
                    tibble(cluster = clustering$pamobject$clustering))
    return(nd)
}

DF <- DF %>% 
    mutate(
        nd = map2(.x = nd,
                   .y = M.no.grade,
                   ~append_pamk_clusters(nd = .x, M = .y))
    )
```

``` {r write_DF, eval = F}
write_rds(DF, 'DF.rds')
```

``` {r read_DF}
DF <- read_rds('DF.rds')
```

By using the Mapper algorithm [reference!] we examine the structure of the data in terms of neighbourhoods of points within the activity space, and how these neighbourhoods are connected. The Mapper algorithm requires the choice of a single variable of the underlying topological space to filtrate on, for which here we choose the *grade* variable. What this means is that the algorithm begins at the low end of the grades and gradually increases the grades whilst looking around and observing how the points in the space cluster and how these clusters connect. 

The most interseting story was comparing the transition of the two subjects from 2017 to 2018 (with the change in resource design), and how this appears to affect the relationship between performance in the subject (*grade*, coloured) and online behaviour in the subject.

``` {r tda_plots}
set.seed(402)
copt = "B"
gridExtra::grid.arrange(
    plot_mapper(1, color_option = copt) + ggtitle("Subject A", "2017"),
    plot_mapper(2, color_option = copt) + ggtitle("Subject A", "2018"),
    plot_mapper(3, color_option = copt) + ggtitle("Subject B", "2017"),
    plot_mapper(4, color_option = copt) + ggtitle("Subject B", "2018")
)
gridExtra::grid.arrange(
    plot_mapper_time(1, color_option = copt) + ggtitle("Subject A", "2017"),
    plot_mapper_time(2, color_option = copt) + ggtitle("Subject A", "2018"),
    plot_mapper_time(3, color_option = copt) + ggtitle("Subject B", "2017"),
    plot_mapper_time(4, color_option = copt) + ggtitle("Subject B", "2018")
)
```

``` {r}
# Spectrums of some of the clusters in the mapper graphs
plot_mapper(1, color_option = "A", use_labels = T) + ggtitle("Subject A", "2017")
plot_timestamp_spectrum(DF$int[[1]], group = mapper_vertex) + 
    facet_wrap(~mapper_vertex) +
    ggtitle("Subject A", "2017")
plot_mapper(2, color_option = "B", use_labels = T) + ggtitle("Subject A", "2018")
plot_timestamp_spectrum(DF$int[[2]], group = mapper_vertex) + 
    facet_wrap(~mapper_vertex) +
    ggtitle("Subject A", "2018")
plot_mapper(3, color_option = "C", use_labels = T) + ggtitle("Subject B", "2017")
plot_timestamp_spectrum(DF$int[[3]], group = mapper_vertex) + 
    facet_wrap(~mapper_vertex) +
    ggtitle("Subject B", "2017")
plot_mapper(4, color_option = "D", use_labels = T) + ggtitle("Subject B", "2018")
plot_timestamp_spectrum(DF$int[[4]], group = mapper_vertex) + 
    facet_wrap(~mapper_vertex) +
    ggtitle("Subject B", "2018")
```

Of note is the 'self regulation' section between 1-10 seconds (wanting this low for higher self regulation), particularly in Sub B 2018.

``` {r sub_B_spect}
# TODO: Colour by grade?
spect_compare_plot <- function(n, clust_var) {
    nodedf <- DF$nd[[n]]
    intdf <- DF$int[[n]]
    if (clust_var %in% names(DF)) {
        sumdf <- mapper_vertex_summaries(DF$Mapper[[n]], nodedf)
        int_aug <- intdf %>% rename(group = mapper_vertex) %>% 
            inner_join(sumdf %>% select(group = id, grade), by = "group")
    } else {
        if (clust_var %in% names(DF$nd[[n]])) {
            int_aug <- intdf 
            int_aug$group <- as.numeric(int_aug[[clust_var]])
            sumdf <- nodedf
            sumdf$group <- sumdf[[clust_var]]
            sumdf <- sumdf %>% 
                group_by(group) %>% 
                summarise(grade = mean(Grade, na.rm = T))
            int_aug <- int_aug %>% 
                inner_join(sumdf, by = "group")
        } else {
            stop(paste(clust_var, "not found in DF or nd data"))
        }
    }
    plot_timestamp_spectrum(int_aug, 
                            group = group,
                            color = grade)
}

spect_compare_plot(4, "cluster_5")
spect_compare_plot(4, "mapper_vertex")

```

### Persistant Homology

It is challenging to compute persistant homology barcodes or birth-death diagrams for 7 or 8 dimensional space, even with bootstrapping small samples, so here I have used the first 3 principal components as a representation for this data. The 3 components exclude grade data, so this is performed again with an additional Grade dimension.

``` {r persistent_homology}
plot(DF$pHomDiag[[1]]$diagram, barcode = T)
plot(DF$pHomDiag[[2]]$diagram, barcode = T)
plot(DF$pHomDiag[[3]]$diagram, barcode = T)
plot(DF$pHomDiag[[4]]$diagram, barcode = T)

plot(DF$pHomDiagGrade[[1]]$diagram, barcode = T)
plot(DF$pHomDiagGrade[[2]]$diagram, barcode = T)
plot(DF$pHomDiagGrade[[3]]$diagram, barcode = T)
plot(DF$pHomDiagGrade[[4]]$diagram, barcode = T)



```